On the link between functional traits and growth rate: meta-analysis shows effects change with plant size, as predicted

### 2016-02-19 MS.tex
Clarify predictions for mass-based growth, plus minor edits throughout

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{King-1999, Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Size has long been known to play a central role in organism design and function \citep{Vogel:1988ux, Vogel:2003wb}, including that of trees \citep{Farnsworth:1995im, Givnish:1995ta, King-2011}. As plants grow larger, a greater amount of energy is devoted towards building and maintaining non-photosynthetic tissues \citep{Givnish:1995ta,King-2011}. A recent model by \citet{Falster:2011ii} suggests this change shifts the benefits and costs of some trait-based trade-offs, such that the net effect of the trait on growth changes with size. Thus the absence of a correlation between SLA and growth at larger sizes may in fact be expected, when potential influences of traits on plant function are considered \citep{Falster:2011ii}. While recent empirical studies support the idea that the relationship between traits and growth rates may change with size (\citealt{Iida:2014ep, Iida:2014hq}), the trend and mechanisms behind this variation remain unclear.

@@

Recent theoretical work \citep{Falster:2011ii} allows us to formulate hypotheses about the expected relationship between traits and growth, based on a mathematical decomposition of plant growth rate. Specific predictions arise from considering the influence of traits on elements of these equations and how these effects may vary with size. This model extends earlier efforts to model growth as a function of traits \citep{Lambers:1992bj,Cornelissen:1996hf,Wright:2000kw,Enquist:2007ek}. However, whereas earlier models focused solely on mass-based relative growth rate and did not predict a change in effect of traits with size, the revised equations do allow for these effects, while also making predictions for both absolute and relative growth in any of mass (eq. \ref{eq:eq1}), height (eq. \ref{eq:eq2}), or diameter (eq. \ref{eq:eq4}).

@@

The \textbf{growth rate in height} can likewise be decomposed as the product of several meaningful terms:
\begin{equation}\label{eq:eq2}
\underbrace{\frac{\ud H}{\ud t}}_{\text{growth rate in height}} = \underbrace{\frac{\ud H}{\ud A_{\rm l}}}_{\text{architecture layout}} \times \underbrace{{\frac{\ud A_{\rm l}}{\ud M_{\rm a} }}}_{\text{leaf deployment per mass}} \times {\frac{\ud M_{\rm a} }{\ud B}} \times {\frac{\ud B}{\ud t}}.
\end{equation}
where $\ud H / \ud A_{\rm l}$, is the growth in plant height
per unit growth in total leaf area. This term expresses the architectural
strategy of the plant. We assume it is unrelated to the traits considered in the current paper, in other words that it does not automatically change in response to changes in the traits. The second term, $\ud A_{\rm l} / \ud M_{\rm a} $,
accounts for the marginal cost of deploying an additional unit of leaf
area, including construction of the leaf itself and various support
structures.  $\ud A_{\rm l} / \ud M_{\rm a} $
can itself be expressed as a sum of construction costs per unit leaf
area,
\begin{equation}\label{eq:daldmt}
\frac{\ud A_{\rm l}}{\ud M_{\rm a} }
= \left(\frac{\ud M_{\rm l}}{\ud A_{\rm l}}
 +  \frac{\ud M_{\rm s}}{\ud A_{\rm l}} + \frac{\ud M_{\rm r}}{\ud A_{\rm l}}\right)^{-1}.
\end{equation}

Similarly \textbf{growth rate in basal area} can be expressed as the sum of growth in sapwood and heartwood, with the former term again decomposed as a product:
\begin{equation}\label{eq:eq3}
\begin{array}{ccccc}
\underbrace{{\frac{\ud A_{\rm st}}{\ud t}}}_{\text{basal area growth rate}} &=& \frac{\ud A_{\rm ss}}{\ud t} &+& \frac{\ud A_{\rm sh}}{\ud t} \\
&=& \underbrace{\frac{\ud A_{\rm ss}}{\ud A_{\rm l}}  \times {\frac{\ud A_{\rm l}}{\ud M_{\rm a} }} \times  {\frac{\ud M_{\rm a} }{\ud B}} \times {\frac{\ud B}{\ud t}}}_{\text{increase in total sapwood}} &+&  \underbrace{k_{\rm ss} \times \frac{\ud A_{\rm ss}}{\ud t}}_{\text{sapwood turnover}},\\
\end{array}
\end{equation}
where $\frac{\ud A_{\rm ss}}{\ud A_{\rm l}} $ is the  sapwood area per leaf area. The \textbf{growth rate in diameter} is then obtained via the geometric relationship between stem diameter and stem area,
\begin{equation}\label{eq:eq4}
\underbrace{\frac{\ud D}{\ud t}}_{\text{growth rate in diameter}} = \sqrt[]{\frac{\Pi}{A_{\rm st}}} \times {\frac{\ud A_{\rm st}}{\ud t}}.
\end{equation}

@@

The above equations are general, i.e. they do not depend on detailed assumptions about plant construction. To make qualitative predictions with respect to traits such as specific leaf area and wood density we require one additional assumption: that the mass of sapwood tissue ($M_{\rm ss}$) increases disproportionately with respect to the plant's leaf area ($A_{\rm l}$). This assumption is included in nearly every model of plant growth that has been developed, on the basis that as stems grow longer the sapwood associated with each leaf must be increasing. It has the effect of changing the relative importance of different components within equations for $\frac{\ud B}{\ud t}$ (net live biomass production) and $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ (marginal cost of deploying additional leaf area) as a plant increases in size. Specifically, in eq. \ref{eq:eq0}, the costs of stem respiration and turnover increase faster than leaf photosynthesis, respiration or turnover, as the plant grows. Similarly, in eq. \ref{eq:daldmt}, the costs of constructing sapwood to support a given unit of leaf area, $\frac{\ud M_{\rm s}}{\ud A_{\rm l}}$, increases faster than the cost of producing a unit of leaf area, $\frac{\ud M_{\rm l}}{\ud A_{\rm l}}$ which is simply the inverse of specific leaf area. Whether plant growth rate increases with a specific trait, therefore depends on how the trait affects the different elements of \ref{eq:eq0}-\ref{eq:eq4} and how those effects change with size (Table \ref{tab:trade-offs}).

Based on the above model and this single assumption about growth, we outline five hypotheses that arise by considering the influence of trait variation on elements of equations  \ref{eq:eq0}-\ref{eq:eq4} and how these may interact with plant size (see Table \ref{tab:trade-offs}). It is possible that other correlations may arise indirectly. Nevertheless, for the purposes of this paper we adopt as working hypotheses only those arising from direct mechanistic causation. We consider predictions for each of height, diameter, and mass-based growth, and how they  may vary depending whether growth is expressed in absolute or relative terms, (see Table \ref{tab:trade-offs}). For many trait and growth combinations, the predicted effect of the trait on growth is similar when expressed in relative and absolute terms. However, this is not true for all traits (details given below).
\subsubsection*{H1: The relationship between SLA and growth rate is expected to shift from positive in seedlings to non-significant or negative in adult plants}

Plants with high SLA can deploy a unit of leaf area more cheaply. However, high-SLA species also have shorter leaf lifespan, and must use more energy replacing leaves lost via turnover \citep{Wright:2004jb}. These two effects occur within the terms  $\frac{\ud B}{\ud t}$ (net live biomass production) and $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ (marginal cost of deploying additional leaf area), appearing on the right side in the equations for growth rate (\ref{eq:eq1}, \ref{eq:eq2}, \ref{eq:eq03}). Whether plant growth rate increases with SLA depends on how the relative magnitude of the benefit (cheap leaf construction, via $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ ) and cost (higher leaf turnover, via $\frac{\ud B}{\ud t}$) of increased SLA changes with size (Table \ref{tab:trade-offs}).
During early life, when leaves comprise a large part of the plant, increasing SLA has an overwhelmingly positive effect on height and diameter growth rate, because the effect of reducing $\frac{\ud M_{\rm l}}{\ud A_{\rm l}}$ (the cost of producing a unit of leaf area) has a large effect compared to the other terms in eq. \ref{eq:daldmt}. As plants increase in size, however, the cost of building sapwood ($\frac{\ud M_{\rm s}}{\ud A_{\rm l}}$) increases disproportionately, decreasing the benefit of cheaper leaf construction in eq.  \ref{eq:daldmt}. Consequently, the relative effect of leaf turnover comes to dominate at larger sizes, and as such, the effect of SLA on height and diameter growth shifts from positive to negative as plants increase in size. A similar shift is expected for growth in mass when expressed in relative terms. However, when expressed as absolute growth, the predicted relationship is negative. This occurs because of the increased turnover of high SLA species.

@@

Assuming plants otherwise maintain a similar cross section of conducting tissue, cheaper stem construction (low wood density or stem density) will increase the marginal rate of leaf deployment per total mass increment ($\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$), and consequently growth rate in height and diameter (eqs. \ref{eq:eq2}-\ref{eq:eq4}; Table \ref{tab:trade-offs}). Disadvantages associated with cheap stem construction take the form of higher mortality and so are not reflected in growth equations. Under this scenario, wood density is negatively correlated to growth rate across all ontogenetic stages. Because the relative expenditure on stem compared to leaf increases as plants grow larger, the relationship between wood density and growth may also strengthen in larger plants. A similar shift is expected for growth in mass when expressed in relative terms. However, when expressed as absolute growth, the predicted relationship is non-significant along all plant size \citep[see also][]{Enquist-1999}.

@@

As plants approach maximum height they increasingly divert energy away from growth and into reproduction (see \citealt{Thomas:1996do,Thomas-2011, Wenk:2015jz}). This effect occurs in the term $\frac{\ud M_{\rm a} }{\ud B}$ (the allocation to vegetative growth) in eqs. (\ref{eq:eq2}-\ref{eq:3}). For seedlings, where all individuals are investing solely in growth, we expect no influence of maximum height on growth (i.e. $\frac{\ud M_{\rm a} }{\ud B} =1$). But once some species begin to produce seed, then species closer to their maximum height value are expected to be allocating more to seed production and less to vegetative growth. Hence when comparing across plants at a given size, species with smaller maximum height are expected to show slower vegetative growth rates. Similar correlations are expected no matter what measure of growth is used (Table \ref{tab:trade-offs}).

@@

The entire dataset included studies differing in the growth measurement used (relative or absolute), the number of species, the experimental conditions (stressed vs unstressed), and the concordance between stages used to measure trait and growth rate. However, these differences are clear criteria for assessing the quality of the data for addressing our hypotheses. Consequently, we adopted a two-stage approach where at first we were conservative by focusing exclusively on relative growth rate, across \underline{\textgreater} 10 species, under unstressed conditions, and where the source of the trait and growth is the same (expect for maximum height and seed mass, where a particular stage is by definition required to measure the trait). This dataset is called the ``ideal'' dataset. Then we compared results with those of a more inclusive dataset - the “entire” dataset- with both relative growth rate and absolute growth rate, and all other criteria relaxed.

@@

\begin{table}[h!]
\centering
\caption{Hypothesised effects of traits on key elements of plant function determining growth rate, as described in eq. \ref{eq:eq0} -- \ref{eq:eq4}. Arrows indicate the effect an increase in trait value would have on each element of the equations, with dashes indicating no effect. Traits are: specific leaf area (= SLA), wood density (= WD), Maximum height (= Hmax), Seed mass, and Maximum photosynthesis per unit leaf area (= Area) For further details, see main text.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{lllcccccc}
  \hline
  & & & {\bf Symbol} & {\bf SLA} & {\bf WD} & {\bf Hmax} & {\bf Seed mass} & {\bf Aarea} \\ \hline
  \\
  \multicolumn{9}{ l } {\textbf{Elements of equations \ref{eq:eq0} -- \ref{eq:eq4}}}  \\
  & \multicolumn{2}{ l } {Leaf deployment per mass}  & $\ud A_{\rm l} / \ud M_{\rm a} $ & \\
    & &     \tabitem leaf  &  & $\uparrow$ & - & - & - & - \\
    & &     \tabitem sapwood & & - & $\downarrow$ & - & - & - \\
    & &     \tabitem root & & - & - & - & - & - \\
  & \multicolumn{2}{ c } {Allocation to vegetative growth} & $\ud M_{\rm a} / \ud B$ & - & - & $\uparrow$ & - & - \\
  & & & & & & \\
  & \multicolumn{2}{ l } {Net live biomass production} & $\ud B / \ud t$ & & & & & \\
  & &     \tabitem photosynthesis & & - & - & - & - & $\uparrow$  \\
  & &     \tabitem respiration   & & - & - & - & - & $\uparrow$  \\
  & &     \tabitem leaf turnover & & $\downarrow$ & - & - & - & - \\
  & \multicolumn{2}{ l } {Architecture} & $\ud H / \ud A_{\rm l}$ & - & - & - & - & - \\
  & \multicolumn{2}{ l } {Sapwood area per leaf} & $ \ud A_{\rm ss} / \ud A_{\rm l}$ & - & - & - & - & - \\
  & \multicolumn{2}{ l } {Sapwood turnover} & $k_{\rm ss} \, A_{\rm ss}$ & - & - & - & - & - \\ \hline
  \\
  \multicolumn{9}{ l } {\textbf{Predicted effect on plant growth rate} (small plant {\sepp} large plant)} \\
  & \multicolumn{2}{ l } {in mass} \\
  & & \tabitem{Absolute} & $\ud M_{\rm a} / \ud t$ & $\downarrow$\sepp$\downarrow$ & $-$\sepp$-$ &  $-${\sepp}$\uparrow$ & $\uparrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & & \tabitem{Relative} & $\ud M_{\rm a} / (\ud t . M_{\rm a} )$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ &  $-${\sepp}$\uparrow$ & $\downarrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & \multicolumn{2}{ l } {in height} \\
  & & \tabitem{Absolute} & $\ud H / \ud t$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\uparrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & & \tabitem{Relative} & $\ud H / (\ud t . H)$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\downarrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & \multicolumn{2}{ l } {in diameter} \\
  & & \tabitem{Absolute} & $\ud D / \ud t$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\uparrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & & \tabitem{Relative} & $\ud D / (\ud t . D)$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\downarrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
\hline
  \end{tabular}
  }
\label{tab:trade-offs}
\end{table}

>>>>

Meanwhile the possibility has been raised that trait-growth correlations may change systematically with size \citep{King-1999, Falster:2011ii, Ruger:2012jv, Iida:2014ep, Iida:2014hq}. Size has long been known to play a central role in organism design and function \citep{Vogel:1988ux, Vogel:2003wb}, including that of trees \citep{Farnsworth:1995im, Givnish:1995ta, King-2011}. As plants grow larger, a greater amount of energy is devoted towards building and maintaining non-photosynthetic tissues \citep{Givnish:1995ta,King-2011}. A recent model by \citet{Falster:2011ii} suggests that increased size shifts the benefits and costs of some trait-based trade-offs, such that the net effect of the trait on growth changes with size. Thus the absence of a correlation between SLA and growth at larger sizes may in fact be expected, when the potential influences of traits on plant function are properly considered \citep{Falster:2011ii}. While some recent empirical studies support the idea that the relationship between traits and growth rates may change with size (\citealt{Iida:2014ep, Iida:2014hq}), the generality of the trend and mechanisms behind this variation remain unclear.

@@

Recent theoretical work \citep{Falster:2011ii} allows us to formulate hypotheses about the expected relationship between traits and growth, based on a mathematical decomposition of plant growth rate. Specific predictions arise from considering the influence of traits on elements of these equations and how these effects may vary with size. This model extends earlier efforts to model mass-based growth of seedlings as a function of traits \citep{Lambers:1992bj,Cornelissen:1996hf,Wright:2000kw,Enquist:2007ek}. However, whereas earlier models focused solely on mass-based relative growth rate and did not predict a change in effect of traits with size, the revised equations do allow for these effects, while also making predictions for both absolute and relative growth in any of mass (eq. \ref{eq:eq1}), height (eq. \ref{eq:eq2}), or diameter (eq. \ref{eq:eq4}).

@@

The \textbf{growth rate in height} can likewise be decomposed as the product of several meaningful terms:
\begin{equation}\label{eq:eq2}
\underbrace{\frac{\ud H}{\ud t}}_{\text{growth rate in height}} = \underbrace{\frac{\ud H}{\ud A_{\rm l}}}_{\text{architecture layout}} \times \underbrace{{\frac{\ud A_{\rm l}}{\ud M_{\rm a} }}}_{\text{leaf deployment per mass}} \times {\frac{\ud M_{\rm a} }{\ud B}} \times {\frac{\ud B}{\ud t}},
\end{equation}
where $\ud H / \ud A_{\rm l}$ is the growth in plant height
per unit growth in total leaf area. This term expresses the architectural
strategy of the plant. We assume it is unrelated to the traits considered in the current paper, in other words that it does not automatically change in response to changes in the traits. The term $\ud A_{\rm l} / \ud M_{\rm a} $
accounts for the marginal cost of deploying an additional unit of leaf
area, including construction of the leaf itself and various support
structures.  $\ud A_{\rm l} / \ud M_{\rm a} $
can itself be expressed as a sum of costs per unit leaf
area for constructing various support tissues,
\begin{equation}\label{eq:daldmt}
\frac{\ud A_{\rm l}}{\ud M_{\rm a} }
= \left(\frac{\ud M_{\rm l}}{\ud A_{\rm l}}
 +  \frac{\ud M_{\rm s}}{\ud A_{\rm l}} + \frac{\ud M_{\rm r}}{\ud A_{\rm l}}\right)^{-1}.
\end{equation}

Similarly \textbf{growth rate in basal area} can be expressed as the sum of growth in sapwood and heartwood, with the former term again decomposed as a product:
\begin{equation}\label{eq:eq3}
\begin{array}{ccccc}
\underbrace{{\frac{\ud A_{\rm st}}{\ud t}}}_{\text{basal area growth rate}} &=& \frac{\ud A_{\rm ss}}{\ud t} &+& \frac{\ud A_{\rm sh}}{\ud t} \\
&=& \underbrace{\frac{\ud A_{\rm ss}}{\ud A_{\rm l}}  \times {\frac{\ud A_{\rm l}}{\ud M_{\rm a} }} \times  {\frac{\ud M_{\rm a} }{\ud B}} \times {\frac{\ud B}{\ud t}}}_{\text{increase in total sapwood}} &+&  \underbrace{k_{\rm ss} \times \frac{\ud A_{\rm ss}}{\ud t}}_{\text{sapwood turnover}}.\\
\end{array}
\end{equation}
Here $\frac{\ud A_{\rm ss}}{\ud A_{\rm l}} $ is the  sapwood area per leaf area. The \textbf{growth rate in diameter} is then obtained via the geometric relationship between stem diameter and stem area,
\begin{equation}\label{eq:eq4}
\underbrace{\frac{\ud D}{\ud t}}_{\text{growth rate in diameter}} = \left(\Pi \, A_{\rm st}\right)^{-0.5} \times {\frac{\ud A_{\rm st}}{\ud t}}.
\end{equation}

@@

The above equations are general, i.e. they do not depend on detailed assumptions about plant construction. To make qualitative predictions with respect to traits such as specific leaf area and wood density we require one additional assumption: that the mass of sapwood tissue ($M_{\rm ss}$) increases disproportionately with respect to the plant's leaf area ($A_{\rm l}$). This assumption is included in nearly every model of plant growth that has been developed, on the basis that as stems grow longer the total mass of sapwood associated with each leaf must be increasing \citep[e.g.][]{Falster:2011ii,King-1999,Enquist:2007ek}. It has the effect of changing the relative importance of different components within equations for $\frac{\ud B}{\ud t}$ (net live biomass production) and $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ (marginal cost of deploying additional leaf area) as a plant increases in size. Specifically, in eq. \ref{eq:eq0}, the costs of stem respiration and turnover increase faster than leaf photosynthesis, respiration or turnover, as the plant grows. Similarly, in eq. \ref{eq:daldmt}, the costs of constructing sapwood to support a given unit of leaf area, $\frac{\ud M_{\rm s}}{\ud A_{\rm l}}$, increases faster than the cost of producing a unit of leaf area, $\frac{\ud M_{\rm l}}{\ud A_{\rm l}}$ which is simply the inverse of specific leaf area. Whether plant growth rate increases with a specific trait, therefore depends on how the trait affects the different elements of \ref{eq:eq0}-\ref{eq:eq4} and how those effects change with size (Table \ref{tab:trade-offs}).

Based on the above model and this single assumption about growth, we outline five hypotheses that arise by considering the influence of trait variation on elements of equations  \ref{eq:eq0}-\ref{eq:eq4} and how these may interact with plant size (see Table \ref{tab:trade-offs}). It is possible that other correlations may arise indirectly. Nevertheless, for the purposes of this paper we adopt as working hypotheses only those arising from direct mechanistic causation. We consider predictions for each of height, diameter, and mass-based growth, and how they  may vary depending whether growth is expressed in absolute or relative terms (see Table \ref{tab:trade-offs}). For many trait and growth combinations, the predicted effect of the trait on growth is similar when expressed in relative and absolute terms. However, this is not true for all traits (details given below).
\subsubsection*{H1: The relationship between SLA and growth rate is expected to shift from positive in seedlings to non-significant or negative in adult plants}

By definition, plants with high SLA can deploy a unit of leaf area more cheaply. However, high-SLA species also have shorter leaf lifespan, and must use more energy replacing leaves lost via turnover \citep{Wright:2004jb}. These two effects occur within the terms  $\frac{\ud B}{\ud t}$ (net live biomass production) and $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ (marginal cost of deploying additional leaf area), appearing on the right side in the equations for growth rate (\ref{eq:eq1}, \ref{eq:eq2}, \ref{eq:eq3}). Whether plant growth rate increases with SLA depends on how the relative magnitude of the benefit (cheap leaf construction, via $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ ) and cost (higher leaf turnover, via $\frac{\ud B}{\ud t}$) of increased SLA changes with size (Table \ref{tab:trade-offs}).
During early life, when leaves comprise a large part of the plant, increasing SLA has an overwhelmingly positive effect on height and diameter growth rate, because the effect of reducing $\frac{\ud M_{\rm l}}{\ud A_{\rm l}}$ (the cost of producing a unit of leaf area) has a large effect compared to the other terms in eq. \ref{eq:daldmt}. As plants increase in size, however, the cost of building sapwood ($\frac{\ud M_{\rm s}}{\ud A_{\rm l}}$) increases disproportionately, decreasing the benefit of cheaper leaf construction in eq.  \ref{eq:daldmt}. Consequently, the effect of leaf turnover becomes relatively stronger at larger sizes, and as such, the effect of SLA on height, diameter and, mass growth shifts from positive to negative as plants increase in size. Note in the case of mass growth, the effect of cheaper leaf construction is realised not in the $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ term , but as a greater total leaf area at a given mass, i.e. the accumulated effect of past reductions in  $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$.

@@

Assuming plants otherwise maintain a similar cross section of conducting tissue, cheaper stem construction (low wood density or stem density) will increase the marginal rate of leaf deployment per total mass increment ($\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$), and consequently growth rate in height and diameter (eqs. \ref{eq:eq2}-\ref{eq:eq4}; Table \ref{tab:trade-offs}). Disadvantages associated with cheap stem construction take the form of higher mortality and so are not reflected in growth equations. Under this scenario, wood density is negatively correlated to growth rate across all ontogenetic stages. Because the relative expenditure on stem compared to leaf increases as plants grow larger, the relationship between wood density and growth may also strengthen in larger plants. A similar shift is expected for growth in mass growth, in either absolute or relative terms. Note in the case of mass growth, the effect of cheaper stem construction is realised not in the $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$ term , but via greater total leaf area at a given mass, i.e. the accumulated effect of past reductions in  $\frac{\ud A_{\rm l}}{\ud M_{\rm a} }$.

@@

As plants approach maximum height they increasingly divert energy away from growth and into reproduction (see \citealt{Thomas:1996do,Thomas-2011, Wenk:2015jz}). This effect occurs in the term $\frac{\ud M_{\rm a} }{\ud B}$ (the allocation to vegetative growth) in eqs. (\ref{eq:eq2}-\ref{eq:eq3}). For seedlings, where all individuals are investing solely in growth, we expect no influence of maximum height on growth (i.e. $\frac{\ud M_{\rm a} }{\ud B} =1$). But once some species begin to produce seed, then species closer to their maximum height value are expected to be allocating more to seed production and less to vegetative growth. Hence when comparing across plants at a given size, species with smaller maximum height are expected to show slower vegetative growth rates. Similar correlations are expected no matter what measure of growth is used (Table \ref{tab:trade-offs}).

@@

The entire dataset included studies differing in the growth measurement used (relative or absolute), the number of species, the experimental conditions (stressed vs unstressed), and the concordance between stages used to measure trait and growth rate. However, these differences are clear criteria for assessing the quality of the data for addressing our hypotheses. Consequently, we adopted a two-stage approach where at first we were conservative by focusing exclusively on relative growth rate, across \underline{\textgreater} 10 species, under unstressed conditions, and where the source of the trait and growth is the same (expect for maximum height and seed mass, where a particular stage is by definition required to measure the trait). This dataset is called the ``ideal'' dataset. We then compared the results from the ``ideal'' dataset with those from the entire dataset - with both relative growth rate and absolute growth rates, and all other criteria relaxed.

@@

\begin{table}[h!]
\centering
\caption{Hypothesised effects of traits on key elements of plant function determining growth rate, as described in eq. \ref{eq:eq0} -- \ref{eq:eq4}. Arrows indicate the effect an increase in trait value would have on each element of the equations, with dashes indicating no effect. Traits are: specific leaf area (= SLA), wood density (= WD), Maximum height (= Hmax), Seed mass, and Maximum photosynthesis per unit leaf area (= Area) For further details, see main text.}
{\footnotesize
\vspace{1cm}
  \begin{tabular}{lllcccccc}
  \hline
  & & & {\bf Symbol} & {\bf SLA} & {\bf WD} & {\bf Hmax} & {\bf Seed mass} & {\bf Aarea} \\ \hline
  \\
  \multicolumn{9}{ l } {\textbf{Elements of equations \ref{eq:eq0} -- \ref{eq:eq4}}}  \\
  & \multicolumn{2}{ l } {Leaf deployment per mass}  & $\ud A_{\rm l} / \ud M_{\rm a} $ & \\
    & &     \tabitem leaf  &  & $\uparrow$ & - & - & - & - \\
    & &     \tabitem sapwood & & - & $\downarrow$ & - & - & - \\
    & &     \tabitem root & & - & - & - & - & - \\
  & \multicolumn{2}{ c } {Allocation to vegetative growth} & $\ud M_{\rm a} / \ud B$ & - & - & $\uparrow$ & - & - \\
  & & & & & & \\
  & \multicolumn{2}{ l } {Net live biomass production} & $\ud B / \ud t$ & & & & & \\
  & &     \tabitem photosynthetic rate & & - & - & - & - & $\uparrow$  \\
  & &     \tabitem respiration rate  & & - & - & - & - & $\uparrow$  \\
  & &     \tabitem leaf turnover rate & & $\downarrow$ & - & - & - & - \\
  & \multicolumn{2}{ l } {Architecture} & $\ud H / \ud A_{\rm l}$ & - & - & - & - & - \\
  & \multicolumn{2}{ l } {Sapwood area per leaf} & $ \ud A_{\rm ss} / \ud A_{\rm l}$ & - & - & - & - & - \\
  & \multicolumn{2}{ l } {Sapwood turnover} & $k_{\rm ss} \, A_{\rm ss}$ & - & - & - & - & - \\ \hline
  \\
  \multicolumn{9}{ l } {\textbf{Predicted effect on plant growth rate} (small plant {\sepp} large plant)} \\
  & \multicolumn{2}{ l } {in mass} \\
  & & \tabitem{Absolute} & $\ud M_{\rm a} / \ud t$ & $\uparrow$\sepp$\downarrow$ & $\downarrow${\sepp}$\downarrow$ &  $-${\sepp}$\uparrow$ & $\uparrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & & \tabitem{Relative} & $\ud M_{\rm a} / (\ud t . M_{\rm a} )$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ &  $-${\sepp}$\uparrow$ & $\downarrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & \multicolumn{2}{ l } {in height} \\
  & & \tabitem{Absolute} & $\ud H / \ud t$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\uparrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & & \tabitem{Relative} & $\ud H / (\ud t . H)$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\downarrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & \multicolumn{2}{ l } {in diameter} \\
  & & \tabitem{Absolute} & $\ud D / \ud t$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\uparrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
  & & \tabitem{Relative} & $\ud D / (\ud t . D)$ & $\uparrow${\sepp}$\downarrow$ & $\downarrow${\sepp}$\downarrow$ & $-${\sepp}$\uparrow$ & $\downarrow${\sepp}$-$ & $\uparrow${\sepp}$\uparrow$ \\
\hline
  \end{tabular}
  }
\label{tab:trade-offs}
\end{table}
